---
title: "6643 HW1"
author: "Guannan Shen"
date: "September 9, 2018"
output: 
  pdf_document:
    latex_engine: lualatex
    number_sections: yes
    toc: yes
    toc_depth: 5
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '5'
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
knitr::opts_chunk$set(engine = "R")
getwd()                                          ## get the path work directory
                                                 ## cache = F, if cache = T, will not revaluate code chunk everytime
## double or more space to insert a line break
```

# Question 2
The data cholesterol.txt contains cholesterol levels (adapted from Rosner, 2006).  The data are a sample of cholesterol levels taken from 24 hospital employees who were on a standard American diet and who agreed to adopt a vegetarian diet for one month.  Serum cholesterol measurements (mcg/dl) were made before adopting the vegetarian diet and one month after.
```{r Q2, echo=TRUE, warning=FALSE}
library(readr)
library(tidyverse)
cholesterol <- read.csv("HW1_cholesterol.csv")
# add the subject id
dim(cholesterol)
cholesterol <- cholesterol %>% mutate(subject_id = seq(1:24))
# add change
cholesterol_w <- cholesterol %>% mutate(change = after - before)
# now model the first 3 models
# change-score model
m1_change_score <- glm(change ~ 1, data = cholesterol_w, family = gaussian)
summary(m1_change_score)

# baseline as covariate model
m2_baseline <- glm(after ~ before, data = cholesterol_w, family = gaussian)
summary(m2_baseline)
# hybrid model
m3_hybrid <- glm(change ~ before, data = cholesterol_w, family = gaussian)
summary(m3_hybrid)
# now make the long form data
cholesterol_l <- cholesterol %>% gather(key = time, value = cholesterol, before:after)

# longitudinal model 
library(lme4)
class(cholesterol_l$time)
lmm_prepost <- lmer(cholesterol ~ time + (1 | subject_id), data = cholesterol_l) 
summary(lmm_prepost)
paste("The AIC of lmm:", AIC(lmm_prepost))



```

# Question 3

```{r Q3, echo=TRUE, warning=FALSE}
library(astsa)
global_temp <- read.csv("HW1_global_temp_anomalies.csv", header = FALSE)
colnames(global_temp) <- c("year", "temp")
# simple linear model
lm_temp <- glm(temp ~ year, data = global_temp, family = gaussian)
summary(lm_temp)
global_temp$res_lm <- resid(lm_temp)
ggplot(global_temp, aes(year, res_lm)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(y = "residuals", subtitle = "residual plot of linear model")
ggsave("Residual_plot_of_linear_model.png", dpi = 600)

## fit ar(1) model for time series data
# using ts() to make a univariate time series object
temp_ts <-  ts(global_temp$temp)
ar_temp <- arima(temp_ts, order = c(1,0,0))
coef(ar_temp)
global_temp$res_ar <- residuals(ar_temp)


## residual plot 
ggplot(global_temp, aes(year, res_ar)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(y = "residuals", subtitle = "residual plot of AR(1) model")
ggsave("Residual_plot_of_AR1_model.png", dpi = 600)

## predict value 
global_temp$y_hat <- global_temp$temp - global_temp$res_ar 
lm_pred <- glm(y_hat ~ year, data=global_temp, family = gaussian)
coef(lm_pred)

# average change per decade
paste("average increase in temp. per decade:", round(10*coef(lm_pred)[2],4))

```
